The Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-08 paper/ms.tex

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to calculate canonical invariants, such as actions, from that measurement. Actions are conserved quantities which describe the orbit of a star under the assumption of regular, bound orbits in a system where the equations of motion are separable in a coordinate system that reflects the system's underlying symmetries. In cylindrical coordinates, which correspond to the axisymmetry commonly assumed for the Galactic disk, the are given by
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the cylindrical radial, vertical, and azimuthal coordinates respectively and the $p_i$ are the conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving
axisymmetric potential, these actions are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

\begin{figure*}
\plotone{fig/midplane_fit.pdf}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial Galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine error bars
(95\uth and 5\uth percentiles), which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

>>>>

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities which describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the typical integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the cylindrical radial, vertical, and azimuthal coordinates respectively and the $p_i$ are the conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving
axisymmetric potential, these actions are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for the toy simulation of a Sagittarius encounter
\citep{2018MNRAS.481..286L}. Again we have subtracted a best fit curve of the
form $\text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}$, with the values
for $\text{A}$, $\text{B}$, and $\text{C}$ being \Gus{something},
\Gus{something}, and \Gus{something else}, respectively. In the first two
panels, we see that the midplane is relatively flat in the inner galaxy. But
additional encounters drive strong midplane variation. In the third panel, we
see a strong $m=2$ mode develop, consistent with the $R=8\,\kpc$ panel of
Figure~17 in \citet{2018MNRAS.481..286L}.

\begin{figure*}
\plotone{fig/midplane_fit.pdf}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine error bars
(95\uth and 5\uth percentiles), which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\plotone{fig/midplane_fit_chervinsim.pdf}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Guessing here, need to
figure out what the panels actually show:}) The left two panels show the
midplane as a function of azimuth before the first encounter. As later
encounters occur, we can see that stronger midplane variations result. In
fact, one can see that an $m=2$ mode is excited in the third panel --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction).}
\label{fig:midplane_chervin}
\end{figure*}
